---
title: '**secrRFS** - SECR random field simulator'
author: "Murray Efford"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_depth: 1
  pdf_document: 
    toc: yes
    toc_depth: 1
vignette: > 
  %\VignetteIndexEntry{SECR random field simulator}
  %\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
header-includes: 
   - \newcommand{\E}[1]{\mbox E (#1)}
   - \renewcommand{\vec}[1]{\mathbf{#1}}
---

\vspace{12pt}

# Introduction

Spatial capture--recapture estimates population density from a sample of $n$ 
marked individuals detected at an array of detectors. Density is treated as 
the intensity of a point process where each point is the activity centre of 
an individual. Overdispersion of $n$ relative to the expected variance 
(Poisson or binomial) leads to poor coverage of confidence intervals for 
density. This document demonstrates the use of **secrRFS** to simulate 
overdispersion of $n$ from a known detection model and generating 
process for the AC. 

The generating process is assumed to be a Cox process i.e. an inhomogeneous 
Poisson process (IHPP) in which the intensity surface varies stochastically 
between realisations (M\o ller and Waagepetersen 2004). The Thomas cluster 
process is treated as a Cox process with intensity surface defined by each 
realisation of parent points. The conditional case (fixed number in known 
area $A$) is not handled.

The available generating processes are

|Generating process|Source function| secrRFS function | Parameters |
|------------------|---------------|------------|-------------|
Binary habitat mosaic | secr::randomHabitat | randomDensity | D, p, A, rescale
Log-gaussian random field | spatstat.random::rLGCP | randomGaussian | D, var, scale
Thomas cluster process | spatstat.random::rThomas | randomParents | D, mu, scale

**secrRFS** requires version 4.6.5 or later of package **secr** (Efford 2025). 
Package spatstat is required for  the log-gaussian and Thomas process options.

# Theory

We define the local density $D_a$ as
$$
D_a \equiv \frac{\int D(\vec x) \, p_\cdot (\vec x | \theta) \, d\vec x}{\int p_\cdot (\vec x | \theta) \, d\vec x}
$$
where $D(\vec x)$ is the intensity at $\vec x$ of one realisation and $p_\cdot(\vec x|\theta)$ is the probability that an AC at $\vec x$ will be detected at least once, given vector of detection parameters $\theta$.

Values of $D_a$ are simulated by generating multiple realisations of $D(\vec x)$.

Overdispersion is estimated as $c = 1 + a(\theta) \mbox{var}(D_a) / \mbox E(D_a).$

# Simple example

See [here](https://www.otago.ac.nz/density/pdfs/secr-overview.pdf) for an overview of 'secr'.

```{r setup, message = FALSE, results = 'hide'}
library(secrRFS)
setNumThreads(2)   # adjust to number of available cores
```

```{r data}
detectpar <- list(lambda0 = 0.5, sigma = 1)
grid144 <- make.grid(12,12, detector = 'proximity', spacing = 2.0)
grid144mask <- make.mask(grid144, spacing = 0.5, buffer = 4)
D <- 256/maskarea(grid144mask)
```

```{r initial, eval = TRUE}
parm1 <- list(D = D, A = 0.5, p = 0.5)
RFS(
	randomfn   = randomDensity,  
	parm       = parm1,
	nrepl      = 1000, 
	traps      = grid144, 
	mask       = grid144mask,
	detectpar  = detectpar,
	noccasions = 5)
```

The result is a value for $c$ that may be used as a variance inflation factor.

# Generating functions 

Parameters are provided in the 'parm' argument whose interpretation is specific 
to each generating function. In each case 'the 'parm' includes the global density D.

## `randomDensity`

This function generates an intensity surface as a random binary habitat mosaic, following Saura and Martinez-Millan (2000) and the `randomHabitat` function in **secr**.

Parameter 'A' (range 0--1) controls the habitat proportion, designated $f$ by Efford and Fletcher (unpubl.).

Parameter 'p' controls fragmentation of the habitat patches. This is non-intuitive. Cluster size increases up to the 'percolation threshold' (about p $\approx 0.59$ in the default case; Saura and Martinez-Millan 2000 p.664).

Parameter 'rescale' (default TRUE) determines whether within-patch density is scale by 1/A to maintain constant global expected density.

```{r plot randomDensity}
Di <- randomDensity(grid144mask,  parm = parm1, plt = TRUE, border = 1, 
	breaks = c(0,10,10000), col=c('grey','white'), legend = FALSE)
pop <- sim.popn(D = Di, grid144mask, model2D = 'IHP')
plot(pop, add = TRUE, frame = FALSE)
```

Fig. 1. Realisation of random habitat mosaic from `randomDensity` (white) with one point simulation.

## `randomGaussian`

This function generates an intensity surface as a log-gaussian random field using rLGCP in the spatstat package.

Parameter 'var' is the variance of *in situ* log intensity. Mean log intensity ($\mu$) is computed as log(D) - var/2.

Parameter 'scale' is the spatial scale of the exponential covariance function.

```{r plot randomGaussian}
parm2 <- list(D = D, var = 0.25, scale = 5)
Di <- randomGaussian(grid144mask,  parm = parm2, plt = TRUE, border = 1)
pop <- sim.popn(D = Di, grid144mask, model2D = 'IHP')
plot(pop, add = TRUE, frame = FALSE)
```

Fig. 2. Realisation of log-gaussian density surface from `randomGaussian` with one point simulation.

## `randomParents`

This function generates an intensity surface from the distribution of parents 
for one realisation of a Thomas process. 

Parameter 'mu' is the expected number of offspring per parent.

Parameter 'scale' is the spatial scale of the circular bivariate normal 
scatter of offspring around the parent.

The surface is derived from the **spatstat** function `rThomas` with argument 'saveLambda = TRUE'.

```{r plot randomParents}
parm3 <- list(D = D, mu = 8, scale = 2)
Di <- randomParents(grid144mask,  parm = parm3, plt = TRUE, parentcex = 0.6,
					border = 5)
pop <- sim.popn(D = Di, grid144mask, model2D = 'IHP')
plot(pop, add = TRUE, frame = FALSE)
```

Fig. 3. Realisation of density surface from `randomParents` with one point simulation. Parent locations are shown as filled dots.

# Parameter combinations

The function `carray` makes it easy to see how varying parameters of the Cox 
process affect overdispersion.

```{r combinations, eval = TRUE}
parmlevels <- list(D = D, mu = 2^(0:5), scale = c(1e-04, 1, 2, 4, 8))
RFSargs <- list (
    randomfn   = randomParents, 
    nrepl      = 1000, 
	traps      = grid144, 
	mask       = grid144mask, 
    detectfn   = 'HHN', 
	detectpar  = detectpar, 
    noccasions = 5)
ca <- carray (parmlevels, RFSargs)
round(ca,1)
```

# Scaling to constant expected number

The parameter list for each generating function includes the component `maskscale`. When this is set to TRUE the intensity surface is scaled to provide a constant expected number within the area of `mask`. The default is `maskscale = FALSE`. For `randomParents` the implied effect is to scale the expected number of offspring per parent $\mu$ to keep $\mbox{E}[N(A)]$ constant.

Scaling can have a large effect on $c$. Here is an example with `randomDensity` (not run).

```{r maskscale, eval = FALSE}
parmlevels <- list(D = D, p = c(0.25,0.5), A = c(0.0625,0.125,0.25,0.5,0.75,1), 
				   maskscale = c(FALSE,TRUE))
RFSargs <- list (
	randomfn   = randomDensity, 
	nrepl      = 1000, 
	traps      = grid144, 
	mask       = grid144mask, 
	detectfn   = 'HHN', 
	detectpar  = detectpar, 
	noccasions = 5)
ca <- carray (parmlevels, RFSargs)
round(ca,1)
```

# References

Baddeley, A. and Turner, R. (2005) spatstat: An R package for analyzing spatial point
  patterns. *Journal of Statistical Software* **12**, 1--42. DOI: 10.18637/jss.v012.i06

Borchers, D. L. and Efford, M. G. (2008) Spatially explicit maximum likelihood methods for capture--recapture studies. *Biometrics* **64**, 377--385.

Efford, M. G. (2023). secr: Spatially explicit capture--recapture models. R package version
  4.6.5. https://CRAN.R-project.org/package=secr/
  

Efford, M. G. and Fletcher, D. J. unpubl. The effect of spatial overdispersion 
on confidence intervals for population density estimated by spatially explicit 
capture--recapture.
  
M\o ller, J., and Waagepetersen, R. P. (2004) Statistical inference and simulation for spatial point processes. Chapman & Hall/CRC.

Saura, S. and Martinez-Millan, J. (2000) Landscape patterns simulation
with a modified random clusters method. *Landscape Ecology*,
**15**, 661--678.

[secr-manual.pdf]: https://www.otago.ac.nz/density/pdfs/secr-manual.pdf
[secr-overview.pdf]: https://www.otago.ac.nz/density/pdfs/secr-overview.pdf
[secr-datainput.pdf]: https://www.otago.ac.nz/density/pdfs/secr-datainput.pdf
[secr-densitysurfaces.pdf]: https://www.otago.ac.nz/density/pdfs/secr-densitysurfaces.pdf
[secr-finitemixtures.pdf]: https://www.otago.ac.nz/density/pdfs/secr-finitemixtures.pdf
[secr-habitatmasks.pdf]: https://www.otago.ac.nz/density/pdfs/secr-habitatmasks.pdf
[secr-markresight.pdf]: https://www.otago.ac.nz/density/pdfs/secr-markresight.pdf
[secr-models.pdf]: https://www.otago.ac.nz/density/pdfs/secr-models.pdf
[secr-multisession.pdf]: https://www.otago.ac.nz/density/pdfs/secr-multisession.pdf
[secr-noneuclidean.pdf]: https://www.otago.ac.nz/density/pdfs/secr-noneuclidean.pdf
[secr-parameterisations.pdf]: https://www.otago.ac.nz/density/pdfs/secr-parameterisations.pdf
[secr-polygondetectors.pdf]: https://www.otago.ac.nz/density/pdfs/secr-polygondetectors.pdf
[secr-sound.pdf]: https://www.otago.ac.nz/density/pdfs/secr-sound.pdf
[secr-spatialdata.pdf]: https://www.otago.ac.nz/density/pdfs/secr-spatialdata.pdf
[secr-telemetry.pdf]: https://www.otago.ac.nz/density/pdfs/secr-telemetry.pdf
[secr-tutorial.pdf]: https://www.otago.ac.nz/density/pdfs/secr-tutorial.pdf
[secr-troubleshooting.pdf]: https://www.otago.ac.nz/density/pdfs/secr-troubleshooting.pdf
[secr-varyingeffort.pdf]: https://www.otago.ac.nz/density/pdfs/secr-varyingeffort.pdf

[secrdesign-vignette.pdf]: https://www.otago.ac.nz/density/pdfs/secrdesign-vignette.pdf
[secrlinear-vignette.pdf]: https://CRAN.R-project.org/package=secrlinear/vignettes/secrlinear-vignette.pdf

[secr-version4.pdf]: https://www.otago.ac.nz/density/pdfs/secr-version4.pdf

[phidot]: http://www.phidot.org/forum/
[secrgroup]: <https://groups.google.com/forum/#!forum/secrgroup>
[CRAN]: https://cran.r-project.org/package=secr
